---
title: "Causes of Deaths"
author: "Deon Vasquez, Ben Ramirez"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing all libraries

```{r}
library("janitor")
library("tidyverse")
library("shiny")
library("shinydashboard")
```
##Loading Data
```{r}
cause_of_death <- read_csv("data/death/cause_of_deaths.csv") %>% clean_names()


```
##Little bit of cleaning to make a happy coder
```{r}
cause_of_death_country <- rename(cause_of_death, country = "country_territory")
cause_of_death_country
```

##Looking at Data Structure
```{r}
glimpse(cause_of_death_country)
```


##Sorting out Non-infectious or Irrelevant modes of death to Travelers.
```{r}
names(cause_of_death_country)
sorted_cause_of_death <- cause_of_death_country%>% 
  select(-c( meningitis, alzheimers_disease_and_other_dementias, parkinsons_disease, nutritional_deficiencies,maternal_disorders, cardiovascular_diseases, neonatal_disorders, self_harm, neoplasms, diabetes_mellitus, chronic_kidney_disease, chronic_respiratory_diseases, cirrhosis_and_other_chronic_liver_diseases,acute_hepatitis))
sorted_cause_of_death
```

##Elongating Data
```{r}
longer_cause_of_death <- sorted_cause_of_death %>%  
  pivot_longer(-c("country","code","year"), 
               names_to = "cause_of_death",
               values_to = "deaths")
longer_cause_of_death
```
##A Little More Cleaning
```{r}
longer_cause_of_death_clean <- longer_cause_of_death %>% 
  rename( year = "year", country_code = "code")
longer_cause_of_death_clean
```

##Selecting Countries of Interest(Asia)/Selecting popular tourist countries in Asia
```{r}
tourist_countries_asia_death <- longer_cause_of_death_clean %>% 
  filter(country_code %in% c("THA","VNM","JPN","IND","IDN","CHN","KHM", "PHL", "MDV","KOR","MYS","MMR","LAO","SGP"))
tourist_countries_asia_death
```
##Sorting into more recent years (2010 and on)
```{r}
modern_cause_of_death <- tourist_countries_asia_death%>% 
filter(year>= 2010)
modern_cause_of_death
```

```{r}


ui <- dashboardPage(
  dashboardHeader(title = "Death in Asia"),
  dashboardSidebar(disable=T),
  dashboardBody(
     fluidRow(
    box(title = "Plot Settings", width =3,   
    selectInput("x", "Select Country", choices = c("Thailand", "Vietnam", "Japan", "India", "Indonesia", "South Korea", "Phillipines", "Laos","Bangladesh", "Maldives", "China","Malaysia", "Myanmar", "Singapore", "Cambodia" ), selected = "Cambodia"),
    selectInput("y", "Select Cause of Death", choices = c("malaria","drowning","interpersonal_violence","hiv_aids","drug_use_disorders","tuberculosis","lower_respiratory_infections","alcohol_use_disorders","exposure_to_forces_of_nature	","diarrheal_diseases","environmental_heat_and_cold_exposure","conflict_and_terrorism","poisonings	","protein_energy_malnutrition","road_injuries","digestive_diseases","fire_heat_and_hot_substances"), selected = "malaria" )
  )
  ),
  box(plotOutput("plot", width = "500px", height = "400px")
)
)
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    modern_cause_of_death %>% 
      filter(country == input$x) %>% 
      filter(cause_of_death==input$y) %>% 
      mutate(year = as_factor(year)) %>% 
      ggplot(aes(x=year, y=deaths, fill=year))+geom_col()+labs(title = "Number of Deaths by Cause per Year from 2010-2019", x= "Year 2010-2019", y="Number of Deaths")
  })
  session$onSessionEnded(stopApp)
}

shinyApp(ui, server)
```

```{r}

south_america %>% 
  
```

```{r}


ui <- dashboardPage(
  dashboardHeader(title = "Death by Asian Country"),
  dashboardSidebar(disable=T),
  dashboardBody(
       
    selectInput("x", "Select Country", choices = c("Thailand", "Vietnam", "Japan", "India", "Indonesia", "South Korea", "Phillipines", "Laos","Bangladesh", "Maldives", "China","Malaysia", "Myanmar", "Singapore", "Cambodia" ), selected = "Thailand"),
  plotOutput("plot", width = "500px", height = "400px")

)
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    modern_cause_of_death %>% 
      
     filter(country == input$x) %>% 
  ggplot(aes(x=year, y= deaths, color = cause_of_death))+
  geom_line()+labs(title = "Deaths yearly sorted by Cause per Country", x= "Year 2010-2019", y="Number of Deaths")
  })
  session$onSessionEnded(stopApp)
}

shinyApp(ui, server)
```
fluidRow(
    box(title = "Plot Settings", width =3, 
box(


```{r}
modern_cause_of_death %>% 
      mutate(year = as_factor(year)) %>% 
     filter(country == "China") %>% 
  ggplot(aes(x=year, y= deaths, color = cause_of_death))+
  geom_line()+labs(title = "Number of Deaths by Cause per Year from 2010-2019", x= "Year 2010-2019", y="Number of Deaths")+scale_y_log10()
```

